Libraries

```{r message=FALSE, warning=FALSE}
library(sf)
library(spatstat)
library(tidyverse)
library(leaflet)
library(spdep)
library(plotly)
library(isdas)
library(gstat)
```

```{r}
#data("Hamilton_CT")
##data("HamiltonDAs") smaller dissemination areas
#Hamilton_CT <- st_set_crs(Hamilton_CT, 26917)
```


```{r}
#Import data
AirData = st_read("C:/Users/Josh/Desktop/4GA3/Applied-Spatial-Statistics-master/PollutionDataTest/AirData/AirData.shx")
summary(AirData)

City = st_read("C:/Users/Josh/Desktop/4GA3/Applied-Spatial-Statistics-master/PollutionDataTest/Hamilton/City_neighbourhoods.shx")
City <- st_transform(City, 26917)
st_crs(City)
```

Plot. Stations are point data, but what we are interested is the concentrations (they are fields)

```{r}
ggplot() +
  geom_sf(data = City, fill = "gray", color = "black", alpha = 0.5, size = 0.3) +
  geom_sf(data = AirData, aes(shape = POLLUTANT_)) +
  coord_sf()
```

I don't know how to deal with temporal data like this, so i chose NO2 pollutants with the enddate of 2022-02-21 as an example.
Going through area data readings:

```{r}
AdataEX = filter(filter(AirData, END_DATE == "2022-02-21"), POLLUTANT_ == "NO2")
```

Leaflet

```{r}
leaflet(data = AdataEX) |>
  setView(lng = -79.86, lat = 43.25, zoom = 10) |>
  addTiles() |>
  addMarkers(data = AdataEX, ~LONGITUDE, ~LATITUDE)
```

Spatially continuous data

```{r}
ggplot() +
  geom_sf(data = City, fill = "gray", color = "black", alpha = 0.5, size = 0.3) +
  geom_sf(data = AdataEX, aes(color = CONCENTRAT, size = CONCENTRAT), alpha = 0.8) +
  scale_color_distiller(palette = "YlOrRd", direction = 1) +
  coord_sf()
```

```{r}
plot_ly(data = AdataEX, x = ~LONGITUDE, y = ~LATITUDE, z = ~CONCENTRAT, marker = list(color = ~CONCENTRAT, colorscale = c("Yellow", "Red"), showscale = TRUE)) |>
  add_markers()
```

Interpolation methods: From lab

```{r}
#Create Voronoi polygons
voronoiP = do.call(c, st_geometry(AdataEX)) |>
  st_voronoi() |>
  st_collection_extract()
voronoiP <- st_set_crs(voronoiP, 26917)

#Replace point with polygon data
poly_data <- AdataEX
poly_data$geometry = voronoiP[unlist(st_intersects(AdataEX, voronoiP))] 

#Bounding box does not work
bbox = st_sf(st_sfc(
  st_polygon(list(rbind(c(-80.3, 43.0), 
                        c(-79.5, 43.0), 
                        c(-79.5, 43.5), 
                        c(-80.3, 43.5), 
                        c(-80.3, 43.0))))
)) #Thank you to the TA for this solution
st_crs(bbox) <- 4326
bbox <- st_transform(bbox, 26917)
poly_data <- poly_data |>
  st_intersection(bbox)

#Plot Voronoi polygons (cannot find a way to )
ggplot(poly_data) +
  geom_sf(aes(fill = CONCENTRAT)) +
  geom_sf(data = AdataEX, size = 0.2) +
  scale_fill_distiller(palette = "YlOrRd", direction = 1) +
  geom_sf(data = City, fill = "gray", color = "black", alpha = 0.1, size = 0.3)

```

IDW

```{r}
#Create owin object
BoxOwin = as.owin(bbox)

#Create ppp object
AdataEX <- relocate(AdataEX, CONCENTRAT, .after = LATITUDE)
Datappp = as.ppp(X = AdataEX[6:8], W = BoxOwin)

#Create IDW surface
idw_map = idw(Datappp, power = 1.5)
plot(idw_map) #Ignore the long and lat maps, I have no idea why its doing that
```

k-point means (I dont even know)

```{r}
#Create prediction grid
Xseq = seq(-80.3, -79.5, 0.005)
Yseq = seq(43.0, 43.5, 0.005)
Target = expand.grid(x = Xseq, y = Yseq) |>
  st_as_sf(coords = c("x", "y"))

#Create k point means
k_data = kpointmean(source_xy = AdataEX, target_xy = Target, z = CONCENTRAT, k = 2) |>
  rename(CONCENTRAT = z)

#Plot
ggplot() +
  geom_sf(data = k_data, aes(color = CONCENTRAT)) +
  scale_color_distiller(palette = "YlOrRd", direction = 1)
```

Trend Surface Estimations
```{r}
AdataEX <- mutate(AdataEX, X3 = LONGITUDE^3, X2Y = LONGITUDE^2 * LATITUDE, X2 = LATITUDE^2, XY = LONGITUDE * LATITUDE, Y2 = LATITUDE^2, XY2 = LONGITUDE * LATITUDE^2, Y3 = LATITUDE^3, X = LONGITUDE, Y = LATITUDE)

#Linear polynomial
Trend1 = lm(formula = CONCENTRAT ~ X + Y, data = AdataEX)
summary(Trend1)

#Quadratic polynomial
Trend2 = lm(formula = CONCENTRAT ~ X2 + X + XY + Y + Y2, data = AdataEX)
summary(Trend2)

#Cubic polynomial
Trend3 = lm(formula = CONCENTRAT ~ X3 + X2Y + X2 + X + XY + Y + Y2 + XY2 + Y3, data = AdataEX)
summary(Trend3)

```

Choosing [trend] because ...:

```{r}
#Interpolation grid
xSeq = seq(-80.3, -79.5, 0.005)
ySeq = seq(43.0, 43.5, 0.005)
Intergrid = expand.grid(X = Xseq, Y = Yseq)
Intergrid <- mutate(Intergrid, X3 = X^3, X2Y = X^2 * Y, X2 = X^2, XY = X * Y, Y2 = Y^2, XY2 = X * Y^2, Y3 = Y^3)
Prediction <- predict(Trend1, newdata = Intergrid, se.fit = TRUE, interval = "prediction", level = 0.95)
summary(Prediction$fit)
```
```{r}
#lm Surface Interpolation
Zp = matrix(data = Prediction$fit[,1], nrow = 101, ncol = 161, byrow = TRUE)
Zp_low = matrix(data = Prediction$fit[,2], nrow = 101, ncol = 161, byrow = TRUE)
Zp_high = matrix(data = Prediction$fit[,3], nrow = 101, ncol = 161, byrow = TRUE)

#Plot 3D
plot_ly(x = ~Xseq, y = ~Yseq, z = ~Zp, type = "surface", colors = "YlOrRd") |>
  add_surface(x = ~Xseq, y = ~Yseq, z = ~Zp_low, opacity = 0.5, showscale = FALSE) |>
  add_surface(x = ~Xseq, y = ~Yseq, z = ~Zp_high, opacity = 0.5, showscale = FALSE) |>
  layout(scene = list(aspectmode = "manual", aspectratio = list(x = 1, y = 1, z = 1)))

#Create add residuals to dataframe
AdataEX$PredRes <- ifelse(Trend1$residuals > 0, "Positive", "Negative")

#Plot residuals
ggplot(data = AdataEX, aes(x = X, y = Y, color = PredRes)) +
  geom_point() +
  coord_equal()
```

Linear looks the best, the others REALLY dont look good
  
Semivariogram 

```{r}
#Get semivariogram of data set
Vari = variogram(CONCENTRAT ~ 1, data = AdataEX)
#Plot
ggplot(data = Vari, aes(x = dist, y = gamma)) + 
  geom_point() + 
  geom_text(aes(label = np), nudge_y = -2) + 
  xlab("Distance") +
  ylab("Semivariance")
```

```{r}
#Semivariogram of the residuals 
AdataEX$TrendRes <- Trend1$residuals
VariRes = variogram(TrendRes ~ 1, data = AdataEX)

#Get best fitting model
VariRes_Th = fit.variogram(VariRes, model = vgm("Exp", "Sph", "Gau"))
VariRes_Th

Fit_line <- variogramLine(VariRes_Th, maxdist = 12500)
ggplot(data = VariRes, aes(x = dist, y = gamma)) + 
  geom_point() + 
  geom_text(aes(label = np), nudge_y = -2) + 
  geom_line(data = Fit_line, aes(x = dist, y = gamma)) +
  xlab("Distance") +
  ylab("Semivariance")
```

Kriging
```{r}
#Kriging 
IntergridSF = Intergrid |> st_as_sf(coords = c("X", "Y"), remove = FALSE, crs = 26917)
IntergridSF <- mutate(IntergridSF, X3 = X^3, X2Y = X^2 * Y, X2 = X^2, XY = X * Y, Y2 = Y^2, XY2 = X * Y^2, Y3 = Y^3) #in case of model change
KrigModel = krige(CONCENTRAT ~ X + Y, filter(AdataEX, OBJECTID != 6), IntergridSF, VariRes_Th) #Bandaid fix, filter out duplicate using object id. Using higher order polynomials still does not work well

#Extract results
Kpred = matrix(data = KrigModel$var1.pred, nrow = 101, ncol = 161, byrow = TRUE) #ISSUE: does not work when there are points in the data with same coordinates

#Plot interpolation
plot_ly(x = ~Xseq, y = ~Yseq, z = ~Kpred, type = "surface", colors = "YlOrRd") |> 
  layout(scene = list(aspectmode = "manual", aspectratio = list(x = 1, y = 1, z = 1)))

```

